A Mechanised Library for Generic (Co)Programming

### 2024-02-23 paper/paper.tex
minor work on sections 1 and 5

This has been a long-standing challenge to come up with a sound algorithm which
rejects all programs that do not terminate (soundness) and accepts as many
well-defined programs as possible (completeness). Our aim is to provide an
additional tool to the every-day programmer to be able to extend the range of
recursive programs we can write.

@@

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.
\mpav{Moved this from Section 3}
goals are: (1) the
mechanisation of hylomorphisms and their laws should be axiom-free; and (2) Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.  Although the use of
\mintinline{OCaml}{Obj.magic} is safe in the code extracted by Coq, it has been
proven that, for higher-order programs, simple interoperations can lead to
incorrect behaviour or even segfaults~\cite{forster:hal-04329663}.  Therefore,
the avoidance of axioms and  \mintinline{OCaml}{Obj.magic} means that the
extracted code will have stronger safety guarantees, and will allow the safe
interoperation with other OCaml code.

@@

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program transformation. This area
has been extensively explored over the years. Tesson et al.\ demonstrated the
efficacy of leveraging Coq to establish an approach for implementing a robust
system dedicated to verifying the correctness of program transformations for
functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and
Emoto went further and\ldots~\cite{10.1007/978-3-030-34175-6_11}

\subsection{Termination of Nonstructural Recursion}

@@

Another approach to certifying termination of nonstructurally recursive
functions is the use of \emph{sized types}. Sized types were introduced by
Hughes et al.\ as a way to track/verify various properties of recursive
programs, including productivity and termination~\cite{HughesPS96}. The core
idea of sized types is that types express bounds on the sizes of recursive data
structures. With this approach, algorithms are implemented in the standard
functional way, following the divide-and-conquer pattern of splitting, recurring,
and merging, with the addition of the data types being indexed by a static
approximation of the relative \emph{size}. An advantage of this approach is
that it allows for code to be written in a natural way. There are costs, however.
Programs must be written in a way that accounts for the handling of size indexes,
and support for the size types must be added to the language.

>>>>

This has been a long-standing challenge to come up with a sound algorithm which
rejects all programs that do not terminate (soundness) and accepts as many
well-defined programs as possible (completeness). Our aim is to provide an
additional tool to the every-day programmer to be able to extend the range of
recursive programs we can write.
\mvol{What is the ``every-day programmer'' here? Are they working in Coq/Agda?}

@@

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.
\mpav{Moved this from Section 3}

The primary goals of this work are:
\begin{enumerate}
  \item that the
  mechanisation of hylomorphisms and their laws should be axiom-free; and
  \item that Coq
hylomorphisms should be extracted to recursive OCaml functions that use no
instances of \mintinline{OCaml}{Obj.magic}.
\end{enumerate}
In OCaml, \mintinline{OCaml}{Obj.magic} allows for unsafe casts between types.
Although the use of \mintinline{OCaml}{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\mintinline{OCaml}{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml code.
\mvol{This is good!}

@@

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs from
their naive counterparts through systematic program transformation. This area
has been extensively explored over the years. Tesson et al.\ demonstrated the
efficacy of leveraging Coq to establish an approach for implementing a robust
system dedicated to verifying the correctness of program transformations for
functions that manipulate lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and
Emoto went further and\ldots~\cite{MurataE19}

\mvol{Need more general citations for program calculation.}

\subsection{Termination Checking}

@@

Another approach to certifying termination of recursive functions is the use of
\emph{sized types}. Sized types were introduced by Hughes et al.\ as a way to
track/verify various properties of recursive programs, including productivity
and termination~\cite{HughesPS96}. The core idea of sized types is that types
express bounds on the sizes of recursive data structures. With this approach,
algorithms are implemented in the standard functional way, following the
divide-and-conquer pattern of splitting, recurring, and merging, with the
addition of the data types being indexed by a static approximation of the
relative \emph{size}. An advantage of this approach is that it allows for code
to be written in a natural way. There are costs, however. Programs must be
written in a way that accounts for the handling of size indexes, and support for
the size types must be added to the language.
